PCA

### 2023-11-08 DRAFT_PCA.Rmd
added .bib, made a bunch of links and added some refs

---
title: "PCA"
authors: "DL Davenport, KE Lotterhos"
date: "2023-11-07"
output:
  html_document:
    df_print: paged
---

@@

This page aims to provide a helpful resource for applying PCA in the field of population genomics. We provide a brief outline of the PCA method and then detail some ‘ins-and-outs’ as it relates to population genomics. Importantly, there are many existing resources for navigating PCA from both the primary literature and online tutorials, several of which we reference throughout this page. These useful resources can be found in the [References](link to section) section at the bottom of this page!

@@

* What is PCA and what are its common uses in population genetics?
* Considerations before sampling:
    + How one chooses to sample from a landscape
    + Sample sizes
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + Batch effects
    + LD and inversion effects
    + Missing data and imputation
    + Hybridization and inbreeding effects
    + Horseshoe and wave patterns
* PCA to inform filtering for other downstream analyses **keep?**
* Comparison to other common approaches to investigating population genetic structure
    + PCA  vs ADMIXTURE/STRUCTURE
* PCA in outlier analysis (i.e. genotype-environment analysis) **keep?**

@@

The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see here for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs). When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called singular value decomposition (SVD) or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e. often more variables than observed entities, n << m, see this youtube playlist for details). Other methods use the covariance (or correlation) matrix. The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (REF Jolliffe, I. (2005). Principal component analysis. Wiley Online Library).

@@

In this respect, PCA is a useful tool to apply to your data, both prior to broader analysis, and as an analysis tool. Below we show how PCA can be applied including how it can be used to check for issues in your data during the filtering/cleaning stage (Batch effects, missing data, linkage disequilibrium, hybridization), how it can be used to highlight informative loci/samples for further exploration (i.e. structural variation, linkage-disequilibrium, sex-linked markers, hybrids), and how it it is commonly used to visualise population structure in a set of unlinked genetic markers.

### Batch effects
Batch effects are caused by technical differences among batches (i.e. groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). These technical differences can result in differences in missing data, genotype error rates, allele frequency estimates, or SNP coverage among batches, causing different batches to appear as unique clusters in PC space. Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). Batch effects can be removed with careful read trimming and filtering (see https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 and references therein).

@@

### LD and Inversion effects
Linkage disequilibrium (LD) is the non-random association between alleles and can arise from physical proximity on a chromosome, demography, or selection. LD is often measured as a correlation in allele frequencies across a set of individuals and varies across a genomic map depending on recombination rates, demographic history, and selection. For example, if two SNPs are physically proximate, the allele at one SNP can predict the allele at the other SNP in the same individual because recombination is rare among sites. But LD can also evolve among SNPs that are not physically proximate, for example if an allele at each SNP is under selection. LD is a source of non-independence in genomic data. Variation in linkage disequilibrium across the map of the genome can warp principal components, as initial PC axes are biased to explain genetic variation in LD as opposed to genetic variation among populations (see Figure 2B above).

@@

### Missing Data in PCA and Imputation

@@

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). Yi and Latch (2021) demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9. Figure 2(b) from Yi and Latch (2021) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares). Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right). As the % of missing data increases, the samples with the most missing data become centered at the PCA origin."}

@@

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10. Figure 4 from Yi and Latch (2021) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population. The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}

@@

### Hybridization and inbreeding effects
Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380 ). However, PC axes can also pick up on differences in homozygosity caused by many generations of inbreeding. These PCA patterns can be informative during early data exploration for identifying cryptic species or hybrids that you may want to remove prior to analyzing for population structure.

@@

In cases of simple two- or three-way admixture, where close relatives of the source populations can be identified and sampled, estimation of admixture proportions can be achieved from projecting samples onto the PCs identified from the source populations. Figure 5 from McVean 2009 below shows an example for human populations from the HapMap3 project. Samples having African ancestry in Southwest USA (ASW) and two groups of samples representing the “source” populations for ASW: (1) Yoruba in Ibadan, Nigeria (YRI) and (2) Utah USA with Northern and Western European ancestry (CEU). The ASW group is admixed between the YRI and CEU groups. The YRI and CEU groups are not necessarily the source populations for the admixed group, but they are closely related to the true source. It is important when using PCA to infer admixture that the PCA is conducted on each chromosome on the two source groups, and all samples are subsequently projected onto the PCA.

@@

| Method or Function, command call | Treatment of Missingness | Ref |
|:---------------------------------|:-------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in (Dray & Josse, 2015); this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see Privé et al., 2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|(Privé et al., 2020; Luu, et al.,  2016)|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |World, 1996|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |**fill in**|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|**fill in**|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (Shriner, 2011).|(Meisner, Liu, et al. 2021;Meisner & Albrechtsen, 2018)|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|**fill in**|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|**fill in**|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|**fill in**|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes |//plink handbook. Randomisation algorithm (aka see)|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in (Wang et al., 2012)|**fill in**|
|smartPCA|Based on Randomized Singular Value Decomposition (RSVD) (Halko et al. 2011), performs full eigendecomposition on the GRM with mean imputation|**fill in**|

>>>>

---
title: "PCA"
bibliography: DRAFT_PCA/PCA_marineomics.bib
authors: "DL Davenport, KE Lotterhos"
date: "2023-11-07"
output:
  html_document:
    df_print: paged
---

@@

This page aims to provide a helpful resource for applying PCA in the field of population genomics. We provide a brief outline of the PCA method and then detail some ‘ins-and-outs’ as it relates to population genomics. Importantly, there are many existing resources for navigating PCA from both the primary literature and online tutorials, several of which we reference throughout this page. These useful resources can be found in the [References](#refs) section at the bottom of this page!

@@

* What is PCA and what are its common uses in population genetics?
* Considerations before sampling:
    + How one chooses to sample from a landscape
    + Sample sizes
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + Batch effects
    + LD and inversion effects
    + Missing data and imputation
    + Hybridization and inbreeding effects
    + Horseshoe and wave patterns
* Comparison to other common approaches to investigating population genetic structure
    + PCA  vs ADMIXTURE/STRUCTURE

@@

The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see [here](https://www.youtube.com/watch?v=PFDu9oVAE-g) for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs). When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called singular value decomposition (SVD) or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e. often more variables than observed entities, n << m, see this [YouTube playlist](https://www.youtube.com/playlist?list=PLMrJAkhIeNNSVjnsviglFoY2nXildDCcv) for details). Other methods use the covariance (or correlation) matrix. The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (@Jolliffe2005-hw).

@@

In this respect, PCA is a useful tool to apply to your data, both prior to broader analysis, and as an analysis tool. Below we show how PCA can be applied including how it can be used to check for issues in your data during the filtering/cleaning stage ([batch effects](#batcheffects), [missing data](#missingdata), [linkage disequilibrium](#ld), [hybridization](#hybrid)), how it can be used to highlight informative loci/samples for further exploration (i.e. structural variation, linkage-disequilibrium, sex-linked markers, hybrids), and how it it is commonly used to visualise population structure in a set of unlinked genetic markers.

### Batch effects {#batcheffects}
Batch effects are caused by technical differences among batches (i.e. groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). These technical differences can result in differences in missing data, genotype error rates, allele frequency estimates, or SNP coverage among batches, causing different batches to appear as unique clusters in PC space. Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). Batch effects can be removed with careful read trimming and filtering (see https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 and references therein).

@@

### LD and Inversion effects {#ld}
Linkage disequilibrium (LD) is the non-random association between alleles and can arise from physical proximity on a chromosome, demography, or selection. LD is often measured as a correlation in allele frequencies across a set of individuals and varies across a genomic map depending on recombination rates, demographic history, and selection. For example, if two SNPs are physically proximate, the allele at one SNP can predict the allele at the other SNP in the same individual because recombination is rare among sites. But LD can also evolve among SNPs that are not physically proximate, for example if an allele at each SNP is under selection. LD is a source of non-independence in genomic data. Variation in linkage disequilibrium across the map of the genome can warp principal components, as initial PC axes are biased to explain genetic variation in LD as opposed to genetic variation among populations (see Figure 2B above).

@@

### Missing Data in PCA and Imputation {#missingdata}

@@

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). Yi and Latch (2022) demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9. Figure 2(b) from Yi and Latch (2022) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares). Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right). As the % of missing data increases, the samples with the most missing data become centered at the PCA origin."}

@@

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10. Figure 4 from Yi and Latch (2022) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population. The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}

@@

### Hybridization and inbreeding effects {#hybrid}
Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380 ). However, PC axes can also pick up on differences in homozygosity caused by many generations of inbreeding. These PCA patterns can be informative during early data exploration for identifying cryptic species or hybrids that you may want to remove prior to analyzing for population structure.

@@

In cases of simple two- or three-way admixture, where close relatives of the source populations can be identified and sampled, estimation of admixture proportions can be achieved from projecting samples onto the PCs identified from the source populations. Figure 5 from McVean 2009 below shows an example for human populations from the [HapMap3 project](https://www.sanger.ac.uk/science/tools/#). Samples having African ancestry in Southwest USA (ASW) and two groups of samples representing the “source” populations for ASW: (1) Yoruba in Ibadan, Nigeria (YRI) and (2) Utah USA with Northern and Western European ancestry (CEU). The ASW group is admixed between the YRI and CEU groups. The YRI and CEU groups are not necessarily the source populations for the admixed group, but they are closely related to the true source. It is important when using PCA to infer admixture that the PCA is conducted on each chromosome on the two source groups, and all samples are subsequently projected onto the PCA.

@@

| Method or Function, command call | Treatment of Missingness | Ref |
|:----------|:------------------------------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in (Dray & Josse, 2015); this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see Privé et al., 2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |@wold1975path|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
 |@purcell2007plink|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|

## References {#refs}
